---
title: "Project3"
author: "Avery Gillespie"
date: "2025-12-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in libraries and UFO csv 
```{r}
library(tidyverse)
library(ggthemes)
library(dplyr)
library(ggplot2)
ufo_csv <- read.csv('~/Downloads/ufo_sightings_scrubbed.csv')
```

Data Cleaning 
```{r}
#using only years after 2000 since it starts in 1949 and changing datetime formatting 
ufo_csv$datetime <- as.character(ufo_csv$datetime)
ufo_csv$datetime <- as.POSIXct(ufo_csv$datetime, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
ufo_csv <- ufo_csv[!is.na(ufo_csv$datetime) & as.numeric(format(ufo_csv$datetime, "%Y")) > 2000, ]

#have only Australia ,Canada ,and USA (major countries represented)
ufo_csv <- ufo_csv[ufo_csv$country %in% c("us", "ca", "au"), ]
```

Import in airport csvs for USA, Australia, and Canada 
```{r}
#us airports csv
us_airports <-read.csv('~/Downloads/airports (4).csv')

#australian airports csv
au_airports <-read.csv('~/Downloads/au-airports.csv')

#canada airports csv
ca_airports <-read.csv('~/Downloads/ca-airports.csv')
```


Include Haversine measurements 
```{r}
haversine <- function(lat1, lon1, lat2, lon2) {
  R <- 6371
  dlat <- (lat2 - lat1) * pi / 180
  dlon <- (lon2 - lon1) * pi / 180
  a <- sin(dlat/2)^2 + cos(lat1*pi/180) * cos(lat2*pi/180) * sin(dlon/2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  R * c
}

nearest_airport_distance <- function(ufo_df, airport_df) {
  distances <- numeric(nrow(ufo_df))
  
  for(i in seq_len(nrow(ufo_df))) {
    lat1 <- ufo_df$latitude[i]
    lon1 <- ufo_df$longitude[i]
    
    dlist <- haversine(lat1, lon1,
                       airport_df$latitude,
                       airport_df$longitude)
    
    distances[i] <- min(dlist, na.rm = TRUE)
  }
  
  distances
}
```


Creating coordinate tables per country 
```{r}
#US airports coordinates table
us_table <- data.frame(
  latitude = us_airports$latitude,
  longitude = us_airports$longitude
)

head(us_table)

#Aussie airports coordinates table
au_table <- data.frame(
  latitude = au_airports$latitude_deg,
  longitude = au_airports$longitude_deg
)

head(au_table)

#canada airports coordinates table 
ca_table <- data.frame(
  latitude = ca_airports$latitude_deg,
  longitude = ca_airports$longitude_deg
)

head(ca_table)




# UFO sightings coordinates table
ufo_table <- data.frame(
  latitude = ufo_csv$latitude,
  longitude = ufo_csv$longitude
)
head(ufo_table)
```


More data cleaning and organizing 
```{r}

#remove na
ufo_table <- ufo_table[!is.na(ufo_table$latitude) & !is.na(ufo_table$longitude), ]

#change to numeric 
ufo_table$latitude <- as.numeric(ufo_table$latitude)
ufo_table$longitude <- as.numeric(ufo_table$longitude)
ufo_table <- ufo_table[!is.na(ufo_table$latitude) & !is.na(ufo_table$longitude), ]



# combined table w all 4 with an extra column for source
ufo_table$source <- "UFO"
us_table$source <- "US Airport"
au_table$source <- "AU Airport"
ca_table$source <- "CA Airport"

combined_table <- rbind(ufo_table, us_table, au_table, ca_table)

head(combined_table)
```

Loading in country shapes and libraries 
```{r}
#load map librarys
library(tidyverse)
library(ggthemes)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# load country shapes
world <- ne_countries(scale = "medium", returnclass = "sf")
```


Prepping for plot
```{r}
#remove Alaska + Hawaii so its bigger 
us_shape <- world %>% 
  filter(admin == "United States of America" & !region_un %in% c("Oceania")) %>%
  st_crop(xmin = -130, xmax = -60, ymin = 20, ymax = 50) 


#more country shapes
ca_shape <- world %>% filter(admin == "Canada")
au_shape <- world %>% filter(admin == "Australia")

# UFO sightings and airport coordinates by country
ufo_us <- ufo_table[ufo_csv$country=="us",]
ufo_ca <- ufo_table[ufo_csv$country=="ca",]
ufo_au <- ufo_table[ufo_csv$country=="au",]

us_airports_coords <- us_table
ca_airports_coords <- ca_table
au_airports_coords <- au_table
```


Creating density map plots per country 
```{r}
# plot density maps
plot_density <- function(country_shape, ufo_coords, airport_coords, country_name) {
  ggplot() +
    geom_sf(data = country_shape, fill = "gray90") +
    stat_density_2d(
      data = ufo_coords,
      aes(x = longitude, y = latitude, fill = after_stat(level)),
      geom = "polygon", alpha = .8
    ) +
    geom_point(
      data = airport_coords,
      aes(x = longitude, y = latitude),
      color = "#FE7F9C", size = .5, alpha = .2
    ) +
    scale_fill_viridis_c(
      option = "viridis",
      name = "Density of Sightings"
    ) +
    theme_minimal() +
    ggtitle(paste("UFO Sightings vs Airports -", country_name)) +
    xlab("Longitude") +
    ylab("Latitude")
}


# US map
plot_density(us_shape, ufo_us, us_airports_coords, "USA")

# Canada map
plot_density(ca_shape, ufo_ca, ca_airports_coords, "Canada")

# Australia map
plot_density(au_shape, ufo_au, au_airports_coords, "Australia")
```

Moving on to other Visualizations ! -----------

Prepping for creating the grouped bar chart 
```{r}
# counts for grouped bar chart
ufo_counts <- ufo_csv %>%
  group_by(country) %>%
  summarize(ufo_sightings = n())

airport_counts <- data.frame(
  country = c("us", "ca", "au"),
  airports = c(nrow(us_table), nrow(ca_table), nrow(au_table))
)

grouped_data <- merge(ufo_counts, airport_counts, by = "country")

#make long
grouped_long <- grouped_data %>%
  pivot_longer(cols = c(ufo_sightings, airports),
               names_to = "type", values_to = "count")
```

Create grouped bar chart 
```{r}
ggplot(grouped_long, aes(x = country, y = count, fill = type)) +
  geom_col(position = "dodge", color = "black", linewidth = 0.3) +
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.9), 
            vjust = -0.3, color = "black", size = 3) +
  scale_x_discrete(labels = c("us" = "USA",
                              "ca" = "Canada",
                              "au" = "Australia")) +
  scale_fill_manual(values = c("ufo_sightings" = "#80EF80",
                               "airports" = "lightblue")) +
  theme_minimal() +
  ggtitle("UFO Sightings vs Number of Airports per Country") +
  xlab("Country") + ylab("Count")
```

Wrangling data for Near vs Far grouped bar chart 
```{r}
haversine <- function(lat1, lon1, lat2, lon2) {
  R <- 6371
  dlat <- (lat2 - lat1) * pi / 180
  dlon <- (lon2 - lon1) * pi / 180
  a <- sin(dlat/2)^2 + cos(lat1*pi/180) * cos(lat2*pi/180) * sin(dlon/2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  R * c
}

nearest_airport_distance <- function(ufo_df, airport_df) {
  distances <- numeric(nrow(ufo_df))
  
  for(i in seq_len(nrow(ufo_df))) {
    lat1 <- ufo_df$latitude[i]
    lon1 <- ufo_df$longitude[i]
    
    dlist <- haversine(lat1, lon1,
                       airport_df$latitude,
                       airport_df$longitude)
    
    distances[i] <- min(dlist, na.rm = TRUE)
  }
  
  distances
}

ufo_us$dist <- nearest_airport_distance(ufo_us, us_table)
ufo_ca$dist <- nearest_airport_distance(ufo_ca, ca_table)
ufo_au$dist <- nearest_airport_distance(ufo_au, au_table)

ufo_dist <- rbind(
  cbind(ufo_us, country = "us"),
  cbind(ufo_ca, country = "ca"),
  cbind(ufo_au, country = "au")
)

ufo_dist$group <- ifelse(ufo_dist$dist <= 50, "Near Airport", "Far Away")
```


Creating grouped bar chart #2
```{r}
ggplot(ufo_dist, aes(x = country, fill = group)) +
  geom_bar(position = "dodge", color = "black", linewidth = 0.3) +
  geom_text(stat = "count", aes(label = ..count..), 
            position = position_dodge(width = 0.9), 
            vjust = -0.3, color = "black", size = 3) +
  scale_fill_manual(
    name = "Distance",
    values = c("Near Airport" = "lightblue",
               "Far Away" = "#C64B8C")
  ) +
  scale_x_discrete(labels = c("us" = "USA",
                              "ca" = "Canada",
                              "au" = "Australia")) +
  theme_minimal() +
  ggtitle("UFO Sightings Near Airports (=<50 km) vs Far Away") +
  xlab("Country") + ylab("Number of Sightings")

#same as above w no usa for better viz 
ufo_dist_no_us <- ufo_dist[ufo_dist$country != "us", ]

ggplot(ufo_dist_no_us, aes(x = country, fill = group)) +
  geom_bar(position = "dodge", color = "black", linewidth = 0.3) +
  geom_text(stat = "count", aes(label = ..count..), 
            position = position_dodge(width = 0.9), 
            vjust = -0.3, color = "black", size = 3) +
  scale_fill_manual(
    name = "Distance",
    values = c("Near Airport" = "lightblue",
               "Far Away" = "#C64B8C")
  ) +
  scale_x_discrete(labels = c("ca" = "Canada",
                              "au" = "Australia")) +
  theme_minimal() +
  ggtitle("UFO Sightings Near Airports (=<50 km) vs Far Away (No USA)") +
  xlab("Country") + ylab("Number of Sightings")
```
